---
title: "Homework 2 Part 1"
author: "Pol Ribó León (1840853)"
date: "25 de junio de 2019"
output: html_document
---

```{r warning=FALSE}
library(invgamma)
library(corrplot)
```


NOTE: Due to problems when knitting, I have been forced to plug-in the variables x and y manually.Nevertheless, the work has been carried out with the standard code for reading files, written below with comments.
```{r}
#data <- read.table("C:/Users/HP/Documents/dugong-data.txt", header = TRUE)

#x <- my_data$Age
#y <- my_data$Length

#Age
x = c( 1.0, 1.5, 1.5, 1.5, 2.5, 4.0, 5.0, 5.0, 7.0, 8.0, 8.5,
       9.0, 9.5, 9.5, 10.0, 12.0, 12.0, 13.0, 13.0, 14.5, 15.5, 
       15.5, 16.5, 17.0, 22.5, 29.0, 31.5)

#Length
y = c(1.80, 1.85, 1.87, 1.77, 2.02, 2.27, 2.15, 2.26, 2.47, 2.19, 
      2.26, 2.40, 2.39, 2.41, 2.50, 2.32, 2.32, 2.43, 2.47, 2.56, 
      2.65, 2.47, 2.64, 2.56, 2.70, 2.72, 2.57)

#number of observations
n <- length(x)

plot(x,y,lwd=2,col="blue",xlab="Age",ylab="Length")

boxplot(x, main="Age")

boxplot(y, main="Length")
```

To highlight, comment that there is only one outlier appearing, in Age, corresponding of an old dugong.

**1b)  Derive the corresponding likelihood function**

\begin{equation}
\begin{aligned} & \prod_{i=1}^{n} \frac{1}{\tau \sqrt{2 \pi}}  \exp \left(-\frac{1}{2}  \frac{\left(Y_{i}-\alpha+\beta \gamma^{x_{i}}\right)^{2}}{\tau^{2}}\right) \\=&\left(\frac{1}{\tau \sqrt{2 \pi}}\right)^{n}  \exp \left(\sum_{i=1}^{n}-\frac{1}{2}  \frac{\left(Y_{i}-\alpha+\beta \gamma^{x_{i}}\right)^{2}}{\tau^{2}}\right) \\=& \frac{1}{(\tau \sqrt{2 \pi})^{n}}  \exp \left(-\frac{1}{2 \tau^{2}}  \sum_{i=1}^{n}\left(Y_{i}-\alpha+\beta \gamma^{x_{i}}\right)^{2}\right) I_{(1, \infty)}(\alpha)I_{(1, \infty)}(\beta)I_{(0,1)}(\gamma)I_{(0, \infty)}(\tau^2)\\ \end{aligned}
\end{equation}


**1c) Write down the expression of the joint prior distribution of the parameters at stake and illustrate your suitable choice for the hyperparameters**

Assuming that the parameters are independent, the joint prior distribution can be found as:

\begin{equation}
\pi(\theta)=\pi(\alpha)  \pi(\beta)  \pi(\gamma)  \pi\left(\tau^{2}\right)
\end{equation}

And the priors for each parameter are:

\begin{equation}
\alpha \quad prior:  
\pi(\alpha)=\frac{1}{\sigma_{\alpha} \sqrt{2 \pi}} \exp \left(-\frac{\alpha^{2}}{2 \sigma_{\alpha}^{2}}\right)
I_{(1, \infty)}(\alpha)
\end{equation}


\begin{equation}
\beta \quad prior:
\pi(\beta)=\frac{1}{\sigma_{\beta} \sqrt{2 \pi}}  \exp \left(-\frac{\beta^{2}}{2 \sigma_{\beta}^{2}}\right)
I_{(1, \infty)}(\beta)
\end{equation}

\begin{equation}
\gamma \quad prior:
\pi(\gamma)=I_{(0,1)}(\gamma)
\end{equation}

\begin{equation}
\tau \quad prior:
\pi\left(\tau^{2}\right)=\frac{b^{a}}{\Gamma(a)} \tau^{2^{(-a-1)}}  \exp \left(\frac{-b}{\tau^{2}}\right)I_{(0, \infty)}\left(\tau^{2}\right)
\end{equation}

Hence, the joint prior distribution is as follows:

\begin{equation}
\pi(\theta)=\frac{1}{\sigma_{\alpha} \sqrt{2 \pi}}  \exp \left(-\frac{\alpha^{2}}{2 \sigma_{\alpha}^{2}}\right)I_{(1, \infty)}(\alpha)  \frac{1}{\sigma_{\beta} \sqrt{2 \pi}}  \exp \left(-\frac{\beta^{2}}{2 \sigma_{\beta}^{2}}\right)I_{(1, \infty)}(\beta)  I_{(0,1)}(\gamma)  \frac{b^{a}}{\Gamma(a)}  \tau^{2^{(-a-1)}}  \exp \left(\frac{-b}{\tau^{2}}\right)I_{(0, \infty)}\left(\tau^{2}\right)
\end{equation}

The choice for the hyperparameters is:

\begin{equation}
\begin{aligned} \sigma_{\alpha} &=10000 \\ \sigma_{\beta} &=10000 \\ a &=0.001 \\ b &=0.001 \end{aligned}
\end{equation}

```{r}
sigma_alpha <- 10000
sigma_beta <- 10000
a <- 0.001
b <- 0.001
```


**1d)  Derive the functional form  (up to proportionality constants) of all full-conditionals**

Given \begin{equation}
\pi\left(\alpha, \beta, \gamma, \tau^{2} | Y\right) \propto \pi\left(Y, \alpha, \beta, \gamma, \tau^{2}\right)
\end{equation}

\begin{equation}
**Full\quad conditional\quad for \quad \alpha**:
\pi\left(\alpha | \beta, \gamma, \tau^{2}, x, y\right)=\frac{1}{\sigma_{\alpha} \sqrt{2 \pi}}  \exp \left(-\frac{\alpha^{2}}{2 \sigma_{\alpha}^{2}}\right)  \frac{1}{(\tau \sqrt{2 \pi})^{n}}  \exp \left(-\frac{1}{2 \tau^{2}}  \sum_{i=1}^{n}\left(Y_{i}-\alpha+\beta \gamma^{x i}\right)^{2}\right)\\
=\frac{1}{\sigma_{\alpha} \sqrt{2 \pi}(\tau \sqrt{2 \pi})^{n}}  \exp \left(-\frac{\alpha^{2}}{2 \sigma_{\alpha}^{2}}-\frac{1}{2 \tau^{2}}  \sum_{i=1}^{n} Y_{i}^{2}-2 Y_{i} \alpha-2 \alpha \beta \gamma^{x_{i}}+\alpha^{2}-2 Y_{i} \beta \gamma^{x_{i}}+\beta^{2} \gamma^{2 x_{i}}\right)\\
 \propto \exp \left(-\frac{\alpha^{2}}{2 \sigma_{\alpha}^{2}}-\frac{1}{2 \tau^{2}}  \sum_{i=1}^{n} \alpha^{2}-2 Y_{i} \alpha-2 \alpha \beta \gamma^{x_{i}}\right)\\
 \propto \exp \left(\frac{\tau^{2}+\sigma_{\alpha}^{2} n}{2 \sigma_{\alpha}^{2} \tau^{2}} \alpha^{2}+\frac{\sigma_{\alpha}^{2} \sum_{i=1}^{n} Y_{i}+\beta \gamma^{x_{i}}}{\sigma_{\alpha}^{2} \tau^{2}} \alpha\right)I_{(1, \infty)}(\alpha)\\
\pi\left(\alpha | \beta, \gamma, \tau^{2}, x, y\right) \sim TN\left(\frac{\sigma_{\alpha}^{2} \sum_{i=1}^{n} Y_{i}+\beta \gamma^{x_{i}}}{\tau^{2}+n \sigma_{\alpha}^{2}}, \frac{\sigma_{\alpha}^{2} \tau^{2}}{\sum_{i=1}^{n} Y_{i}+\beta \gamma^{x_{i}}}\right)
\end{equation}

Note: TN=Truncated Normal distribution as its domain is "truncated". (1,+$\inf$)


\begin{equation}
**Full\quad conditional\quad for \quad \beta**:
\pi\left(\beta | \alpha, \gamma, \tau^{2}, x, y\right)
=\frac{1}{\sigma_{\beta} \sqrt{2 \pi}} \exp \left(-\frac{\beta^{2}}{2 \sigma_{\beta}^{2}}\right)  \frac{1}{(\tau \sqrt{2 \pi})^{n}}  \exp \left(-\frac{1}{2 \tau^{2}}  \sum_{i=1}^{n}\left(Y_{i}-\alpha+\beta \gamma^{x_{i}}\right)^{2}\right)\\
=\frac{1}{\sigma_{\beta} \sqrt{2 \pi}(\tau \sqrt{2 \pi})^{n}}  \exp \left(-\frac{\beta^{2}}{2 \sigma_{\beta}^{2}}-\frac{1}{2 \tau^{2}}  \sum_{i=1}^{n} Y_{i}^{2}-2 Y_{i} \alpha-2 \alpha \beta \gamma^{x_{i}}+\alpha-2 Y_{i} \beta \gamma^{x_{i}}+\beta^{2} \gamma^{2 x_{i}}\right)\\
\propto \exp \left(-\frac{\beta^{2}}{2 \sigma_{\beta}^{2}}-\frac{1}{2 \tau^{2}}  \sum_{i=1}^{n} \beta^{2} \gamma^{2 x_{i}}-2 \alpha \beta \gamma^{x_{i}}-2 Y_{i} \beta \gamma^{x_{i}}\right)\\
\propto \exp \left(-\frac{\beta^{2}\left(\tau^{2}-\sigma_{\beta}^{2} \sum_{i=1}^{n} \gamma^{2 x_{i}}\right)+2 \sigma_{\alpha}^{2} \sum_{i=1}^{n}-\alpha \gamma^{x_{i}}-Y_{i} \gamma^{x_{i}}}{2 \sigma_{\beta}^{2} \tau^{2}}\right)\\
\propto \exp \left(\frac{\tau^{2}+\sigma_{\beta}^{2} \sum_{i=1}^{n} \gamma^{2 x_{i}}}{2 \sigma_{\beta}^{2} \tau^{2}} \beta^{2}+\frac{\sigma_{\beta}^{2} \sum_{i=1}^{n} \gamma^{x_{i}}\left(-\alpha-Y_{i}\right)}{\sigma_{\beta}^{2} \tau^{2}} \beta\right)I_{(1, \infty)}(\beta)
\pi\left(\beta | \alpha, \gamma, \tau^{2}, x, y\right)\\
\sim T N\left(\frac{\sigma_{\beta}^{2} \sum_{i=1}^{n} \gamma^{x_{i}}\left(-\alpha-Y_{i}\right)}{\tau^{2}+\sigma_{\beta}^{2} \sum_{i=1}^{n} \gamma^{2 x_{i}}}, \frac{\sigma_{\beta}^{2} \tau^{2}}{\tau^{2}+\sigma_{\beta}^{2} \sum_{i=1}^{n} \gamma^{2 x_{i}}}\right)
\end{equation}



Note: TN=Truncated Normal distribution as its domain is "truncated". (1,+$\inf$)\\


\begin{equation}
**Full\quad conditional\quad for \quad \gamma**:
\pi\left(\gamma | \alpha, \beta, \tau^{2}, x, y\right)= \frac{1}{(\tau \sqrt{2 \pi})^{n}}  \exp \left(-\frac{1}{2 \tau^{2}}  \sum_{i=1}^{n}\left(Y_{i}-\alpha+\beta \gamma^{x_{i}}\right)^{2}\right)
I_{(0,1)}(\gamma)
\end{equation}


\begin{equation}
**Full\quad conditional\quad for \quad \tau^{2}**:
\pi\left(\tau^{2} | \alpha, \beta, \gamma, x, y\right)=\frac{b^{a}}{\Gamma(a)}  \tau^{2^{(a-1)}}  \exp \left(\frac{-b}{\tau^{2}}\right)  \frac{1}{(\tau \sqrt{2 \pi})^{n}}  \exp \left(-\frac{1}{2 \tau^{2}}  \sum_{i=1}^{n}\left(Y_{i}-\alpha+\beta \gamma^{x_{i}}\right)^{2}\right)\\
\propto \frac{1}{\tau^{2\left(a+1+\frac{n}{2}\right)}} \exp \left(-\frac{b+\frac{1}{2} \sum_{i=1}^{n}\left(Y_{i}-\alpha+\beta \gamma^{x_{i}}\right)^{2}}{\tau^{2}}\right)
I_{(0, \infty)}(\tau)\\
\pi\left(\tau^{2} | \alpha, \beta, \gamma, x, y\right) \sim I Gamma\left(a+\frac{n}{2}, b+\frac{1}{2} \sum_{i=1}^{n}\left(Y_{i}-\alpha+\beta \gamma^{x_{i}}\right)^{2}\right)
\end{equation}

**1e)  Which distribution can you recognize within standard parametric families so that direct simulation from full conditional can be easily implemented ?**

Except for $\gamma$, the other parameters $\alpha$, $\beta$ and $\tau^{2}$ can be recognized as Truncated Normal, Truncated Normal and Inverse Gamma respectively.

**1f) Using a suitable Metropolis-within-Gibbs algorithm simulate a Markov chain (T = 10000) to approximate the posterior distribution for the above model**

```{r}
# Full conditional for alpha
fc_alpha= function(beta, gamma,tau_2){
  mu=sigma_alpha*sum(y+beta*gamma^x)/(n*sigma_alpha+tau_2)
  sd=(tau_2*sigma_alpha)/(n*sigma_alpha + tau_2)
  alpha1=0
  if(alpha1 < 1) {
    alpha1=rnorm(1,mu,sqrt(sd))
  }else{
  return(alpha1)
  }
}  


# Full conditional for beta
fc_beta= function(alpha,gamma,tau_2){
  mub=(sigma_beta*sum((alpha-y)*gamma^x))/(sigma_beta*sum(alpha- y)*gamma^(2*x) + tau_2)
  sdb=sqrt((tau_2*sigma_beta)/(sigma_beta*sum(gamma^(2*x)) +tau_2))
  beta1=0
  if(beta1< 1){
    beta1=rnorm(1,mub,sdb)
  }
    else{
      return(beta1)
  }
  
}  

# Full conditional for gamma
fc_gamma <- function(gamma, alpha, beta, tau_2){
   return(exp(-1/(2*tau_2)*sum((y-alpha+beta*gamma^x)^2)))
}

# Full conditional for tau^2
fc_tau_2 <- function(alpha, beta, gamma){
  
  mut <- n/2 + a
  mut2=b+1/2*sum((y-alpha+beta*gamma^x)^2)
  tau_21 <- rinvgamma(1, mut, mut2)

  return(tau_21)
}

```


```{r}
#SIMULATION


M = 10000

# parameters
alpha_vec  <-rep(NA,M+1)
beta_vec   <- rep(NA,M+1)
gamma_vec  <- rep(NA,M+1)
tau_2_vec  <- rep(NA,M+1)

alpha_vec[1]  <- 1
beta_vec[1] <- 1
gamma_vec[1]  <- 0.5
tau_2_vec[1]  <- 0.5

#hyperparameters
sigma_alpha <- 10000
sigma_beta <- 10000
a <- 0.001
b <- 0.001

for (i in 1:M){
  alpha_vec[i+1] <- fc_alpha(beta = beta_vec[i], 
                             gamma = gamma_vec[i], 
                             tau_2 = tau_2_vec[i])
    
  


  beta_vec[i+1] <- fc_beta(alpha = alpha_vec[i+1], 
                           gamma = gamma_vec[i],
                           tau_2 = tau_2_vec[i])

  gamma_vec[i+1] <- fc_gamma(gamma=gamma_vec[i], 
                             alpha = alpha_vec[i+1],
                            beta = beta_vec[i+1], 
                             tau_2 = tau_2_vec[i])

  
  tau_2_vec[i+1] <- fc_tau_2(alpha = alpha_vec[i+1], 
                             beta = beta_vec[i+1],
                            gamma = gamma_vec[i+1])
    
   
  
      }
```

```{r}
head(cbind(alpha_vec, beta_vec, gamma_vec, tau_2_vec),10)

```


**1g)  Show the 4 univariate trace-plots of the simulations of each parameter**
```{r}
plot(alpha_vec,type='l',main='alpha')
plot(beta_vec,type='l',main='beta')
plot(gamma_vec,type='l',main='gamma')
plot(tau_2_vec,type='l',main='tau_2')
```

**1h)  Evaluate graphically the behaviour of the empirical averages $\hat{I}_t$  with growing $t=1,...,T$**
```{r}

# alpha 
plot(cumsum(alpha_vec)/(1:length(alpha_vec)), type="l",ylim = c(2, 2.8), ylab="", main="Behaviour EA Alpha", xlab="Nº simulations",col='green',lwd=2) 


# beta
plot(cumsum(beta_vec)/(1:length(beta_vec)), type="l",ylim = c(0,100), main="Behaviour EA Beta", xlab="Nº simulations",col='green',lwd=2) 


# gamma
plot(cumsum(gamma_vec)/(1:length(gamma_vec)), type="l",ylim = c(0,.08), main="Behaviour EA Gamma", xlab="Nº simulations",col='green',lwd=2) 


#tau_2
plot(cumsum(tau_2_vec)/(1:length(tau_2_vec)), type="l",ylim = c(0,.8), main="Behaviour EA Tau_2", xlab="Nº simulations",col='green',lwd=2) 

```

**1i)  Provide estimates for each parameter together with the approximation error and explain how you have evaluated such error**

The estimate that it will be used it will be the mean of each parameter.

```{r}
#Estimates

cat("Estimate alpha:", mean(alpha_vec),"\nEstimate beta :", mean(beta_vec),"\nEstimate gamma", mean(gamma_vec), "\nEstimate tau^2", mean(tau_2_vec))
```

To find the approximation error, we will compute the variance over the length of the Markov chain.

```{r}

cat("Approximation error alpha:", var(alpha_vec)/length(alpha_vec),"\nApproximation error beta :", var(beta_vec)/length(beta_vec),"\nEstimatApproximation error gamma", var(gamma_vec)/length(gamma_vec), "\nApproximation error tau^2", var(tau_2_vec)/length(tau_2_vec))
```

**1l)  Which parameter has the largest posterior uncertainty? How did you measure it?**

The largest posterior uncertainty can be measured checking the standard error of each parameter towards its estimate(the mean of the parameter).Hence,

```{r}
cat("Approximation error alpha:", sd(alpha_vec)/mean(alpha_vec),"\nApproximation error beta :", sd(beta_vec)/mean(beta_vec),"\nEstimatApproximation error gamma", sd(gamma_vec)/mean(gamma_vec), "\nApproximation error tau^2", sd(tau_2_vec)/mean(tau_2_vec))

```

Beta has the largest uncertainty, most likely because at the early stage of the simulation its value fluctuates too much. 


**1m)  Which couple of parameters has the largest correlation (in absolute value)?**

```{r}
cor(cbind(alpha_vec, beta_vec, gamma_vec, tau_2_vec))
```

```{r}
corr = cor(data.frame(alpha_vec, beta_vec, gamma_vec, tau_2_vec))
corrplot.mixed(corr)

```
As it can be appreciated, though there aren't high coefficient values, alpha and gamma are the most correlated


**1n)  Use the Markov chain to approximate the posterior predictive distribution of the length of a dugong with age of 20 years.**

```{r}
dug_20 = rep(NA, 10000)

for(i in 1:M){
  mu=alpha_vec[i]+beta_vec[i]*gamma_vec[i]^20
  sd=sqrt(tau_2_vec[i])  
  dug_20[i]= rnorm(1, mu,sd)
  
}

cat("prediction", mean(dug_20))
```


**1o)  Provide the prediction of a different dugong with age 30** 

```{r}
dug_30 = rep(NA, 10000)

for(i in 1:M){
  mu=alpha_vec[i]+beta_vec[i]*gamma_vec[i]^30
  sd=sqrt(tau_2_vec[i])  
  dug_30[i]= rnorm(1, mu,sd)
  
}

cat("prediction", mean(dug_30))
```

**CONCLUSION**: this incredible little difference between predictions could be due to the fact that once a dugong reaches an age close to 20, they don't grow more.

**1p)  Which prediction is less precise?**

We will take a look at the approximation error and the uncertainty.
```{r}

#Approximation error
err_dug_20 = var(dug_20)/length(dug_20)
err_dug_30 = var(dug_30)/length(dug_30)

err_dug_20
err_dug_30
```

```{r}

#Uncertainty
un_dug_20 = sd(dug_20)/mean(dug_20)
un_dug_30 = sd(dug_30)/mean(dug_30)

un_dug_20
un_dug_30
```

As it can be seen,althoug for a very little values, both error and uncertainty of dugongs of age 30 are higher, so we can conclude that the prediction for age 20 is a bit less precise. 
