---
title: "HW1 part 2"
author: "Pol Ribo Leon; 1840853"
date: "15 de mayo de 2019"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
To find the marginal likelihood we assume a model for the data, which depends on a variable theta. As we are not sure on the exact value of theta, we choose it to follow a distribution p(theta|alpha) specified by a constant parameter alpha. So, now the distribution of X is directly determined by alpha . This way, we can calculate the marginal likelihood by integrating out the variable theta. 

As of this;
Marginal Likelihood
$m(y)=\int_{\Theta}L(\theta)\pi(\theta)d\theta$,

Given our likelihood function of the chosen model(Poisson,2)

$L(\theta)= \prod_{i=1}^n f(yi|\theta)= \prod_{i=1}^n \frac{\exp^{-\theta}\theta^{yi}}{yi!}= \frac{e^{-\theta n} \theta^{\sum_{i=1}^{n} yi}} {\prod_{i=1}^n yi!}$

We choose theta to follow a Gamma(1,1) distribution.
$\theta\approx Gamma(1,1)$. 

As, so $\pi(\theta)= \frac{\beta^\alpha}{\Gamma(\alpha)} \theta^{\alpha-1}e^{-\beta\theta}$-> $\pi(\theta)=e^{-\theta}$

We integrate, over the support of the Gamma distribution->(0,Inf)

**Solving the integral**

$m(y)=\int_0^\infty \frac{e^{-\theta n} \theta^{\sum_{i=1}^{n} yi}} {\prod_{i=1}^n yi!}e^{-\theta}d\theta= \frac{1}{{\prod_{i=1}^n yi!}}\int_0^\infty e^{-\theta(n+1)}\theta^{\sum_{i=1}^{n} yi}d\theta$

Now we take advantage of the property distribution support is always equal 1.

$\int_0^\infty \frac{\beta^\alpha}{\Gamma(\alpha)}\theta^{\alpha-1}e^{-\beta\theta}d\theta= 1$ 

$\int_0^\infty \theta^{\alpha-1}e^{-\beta\theta}d\theta= \frac{\Gamma(\alpha)}{\beta^\alpha}$

This way $\alpha=\sum_{i=1}^nyi+1$ and $\beta=n+1$

Then, by substitution, the computation follows;

$\int_0^x \theta^{\sum_{i=1}^n y}e^{-(n+1)\theta}d\theta=\frac{\Gamma(\sum_{i=1}^{n} yi+1)}{(n+1)^{\sum_{i=1}^nyi+1}}$

And finally,

$m(y1,...yn)=\int_0^x \theta^{\sum_{i=1}^n y}e^{-(n+1)\theta}d\theta=\frac{\Gamma(\sum_i=1^n yi+1)}{\prod_{i=1}^{n}yi!(n+1)^{\sum_{i=1}^nyi+1}}=\frac{\Gamma(25)}{1327104(11^{25})}=4.315028e-09$

```{r}

set.seed(123)

pois=rpois(10,2)
prod=1

for (i in 1:10){
  print(pois[i])
  prod=prod*factorial(pois[i])
}

#Gamma(25)=!24
factorial(24)/(prod*(11^25))
```

